---
title: "Practical Meta-analysis in R"
author:
- "Saskia le Cessie"
- "Katerina Papadimitropoulou"
date: "5 April 2018"
output: beamer_presentation
fontsize: 10pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## R and R-studio 

- R is a programming language, originally 'written' in 1992 by Ross Ihaka and Robert Gentleman (yes, "R" stems from the first letter of their first names), inspired by another programming language called S.

- Throughout this practical you'll be encouraged to use R-studio. R-studio is a free, open-source "Integrated Development Environment" (IDE) for R (for Windows, Mac and Linux).

-  Installing R-studio requires only two steps:
     +  Download and install "base R" from the R project website: <https://cloud.r-project.org/>.
     +  Download and install R-studio from the R-studio website <https://www.rstudio.com/products/rstudio/download/>.

## Install and load libraries

```{r, eval=FALSE, echo=TRUE}

# load libraries
install.packages(c("foreign", "metafor", "meta"))

library(foreign)
library(metafor)
library(meta)

# set your working directory
setwd("D:/Meta-analysis course LUMC")
```



## EXERCISE I - Diuretic dataset

```{r, eval=FALSE, echo=TRUE}
# Load the dataset 
diuretic = read.dta("diuretic.dta")

# Take a look at the data 
View(diuretic)
```

- There are two ways to calculate the (log) odds ratios 
     +  Manually; fist calculating the number of healthy subjects (without pre-eclampsia)
     +  Using "escalc" function in R 
     
```{r, message=F, warning=F}
load(file="diuretic.RData")
## load libraries
library(foreign)
library(metafor)
library(meta)
```

```{r,echo=FALSE,message=FALSE,warning=FALSE}
require(knitr)
# Set so that long lines in R will be wrapped:
opts_chunk$set(tidy.opts=list(width.cutoff=50),tidy=TRUE)
```


## EXERCISE I - Perform the MA 

```{r, echo = TRUE}
res1 = rma(yi, vi, data=diuretic, method="FE", measure="OR")
res1
```

## Forestplot - Fixed effects analysis

```{r diuretic}
hc = with(diuretic,nc-pec)
ht = with(diuretic,nt-pet)

forest(res1, xlim=c(-16, 6), at=log(c(0.05, 0.25, 1, 4)), order = "obs", atransf=exp,
       slab = diuretic$trialid, 
       ilab = cbind(diuretic$pet, ht, diuretic$pec, hc),
       ilab.xpos=c(-9.5,-8,-6,-4.5), cex=0.75, ylim=c(-1,12),
       refline = res1$b)

op <- par(cex=1, font=4)

## switch to bold font
par(font=2)

### add column headings to the plot
text(c(-9.5,-8,-6,-4.5), 11, c("pet+", "ht", "pec+", "hc"))
text(c(-8.75,-5.25),     12, c("Treated", "Control"))
text(-16,                11, "Study",  pos=4)
text(6,                  11, "Odds Ratio [95% CI]", pos=2)


## set par back to the original settings
par(op)
```

## Random effects meta-analysis

```{r, eval=FALSE, echo=TRUE}
res2 = rma(yi, vi, data=diuretic, method="REML", measure="OR") # default 
```

Many iterarive and non-iterative approaches to estimate $\bold{\tau^2}$:

\[
\begin{aligned}
Y_i & =\theta + \theta_{i} +\epsilon_{i} \\
\theta_{i} & \sim N(0,\tau^2) \\
\epsilon_{i} & \sim N(0, V_{i})
\end{aligned}
\]
 
 - DL: DerSimonian-Laird (most commonly used)
 - EB: Empirical Bayes 
 - HE: Hedges
 - Hs: Hunter-Schmidt
 - ML: Maximum-likelihood 
 - REML: Restricted maximum-likelihood (default)
 - SJ: Sidik-Jonkman

## EXERCISE II - Streptokinase dataset 
```{r, message=F, warning=F}
load(file="strep.RData")
```
Odds ratio as outcome - using function "metabin" from package meta

```{r }
h1 = with(strep,pop1-cases1)
h0 = with(strep,pop0-cases0)

res1 = metabin(cases1, h1, cases0, h0, data = strep, studlab = trialnam, sm="OR", method="MH", comb.random=TRUE)
forest(res1,  order = "obs", atransf=exp,
       slab = strep$trialnam, 
       refline = res1$b)
```

##  Cumulative meta-analysis 

```{r,  echo=TRUE}
res4 = metacum(res1, sortvar=year, pooled="random")
forest(res4) 
```

## EXERCISE III - cholesterol dataset 

```{r, message=F, warning=F}
load(file="chol.RData")
```
Fixed effects meta-analysis - Forestplot
```{r}
res1 = rma(yi=logor, vi=varlogor, data=chol , method="FE", measure="OR")
forest(res1, atransf=exp,  slab = chol$trialname)
```

## Random effects meta-analysis 
```{r}
res2 = rma(yi=logor, vi=varlogor, data=chol , method="REML", measure="OR")
res2
```

## Forestplot with prediction interval 

```{r,  echo=TRUE}
forest(res2,  atransf=exp,
       slab = chol$trialname, addcred=TRUE, col=c("blue", "red"))
```

## Meta-regression
 Cholesterol level reduction as covariate 

```{r,  echo=FALSE}
res4 = rma(y=logor, vi=varlogor, mods = ~ cholreduc , data=chol)

forest(res4,  atransf=exp,
       slab = chol$trialname)
```

## Meta-analytic Scatterplot
Observed outcomes (ORs) vs individual studies plotted against cholreduc
```{r,  echo=FALSE}
wi = 1/sqrt(chol$varlogor)
size = 0.5 + 3.0 * (wi - min(wi))/(max(wi) - min(wi))

preds = predict(res4, transf=exp, newmods=c(0:2), digits=4)


### plot the odds ratios against absolute latitude
plot(chol$cholreduc, exp(chol$logor), pch=19, cex=size, 
     xlab="Cholesterol reduction", ylab="Odds Ratio",
     las=1, bty="l", log="y")

### add predicted values (and corresponding CI bounds)
lines(0:2, preds$pred)
lines(0:2, preds$ci.lb, lty="dashed")
lines(0:2, preds$ci.ub, lty="dashed")

### dotted line at OR=1 (no difference between groups)
abline(h=1, lty="dotted")


### dotted line at OR=1 (no difference between groups)
abline(h=1, lty="dotted")

ids <- c(12,22,23,28)
pos <- c(3,3,1,1)
text(chol$cholreduc[ids], exp(chol$logor)[ids], ids, cex=0.9, pos=pos)
```

## Funnelplot
```{r,  echo=FALSE}
# Fit first a RE model without any covariates
res2 = rma(yi=logor, vi=varlogor, data=chol , method="REML", measure="OR")

### set up 2x2 array for plotting
par(mfrow=c(2,2))

# draw funnel plots
funnel(res2, main="Standard Error")
funnel(res2, yaxis="vi", main="Sampling Variance")
funnel(res2, yaxis="seinv", main="Inverse Standard Error")
funnel(res2, yaxis="vinv", main="Inverse Sampling Variance")
```
